### 2024-01-15 iterative_monomorphisation.tex
continued expanding on draft focusing mostly on the detailed description of the algorihtm

TODO:
    \begin{itemize}
       \item \st{contact Sascha Boehme about authorship}
        \item are there any implementation of this (or comparable) algorithm in that have been evaluated?
        \item finish implementation (mostly there)
        \item test and evaluate implementation (well on the way)
        \item (can wait) merge implementation (review code with active contributor(s) beforehand)
        \item find pretty inline setup for pseudo code
        \item (in order) write high/low level algo
        \item write implementation and evaluation
        \item write background
        \item write related work
        \item write conclusion
        \item write intro
        \item reread, spellcheck ...
    \end{itemize}

@@

@@

This algorithm is incomplete, this is because not all possible instantiations of the type variables are generated.
Consider the following problem: 
\begin{align*}
   P = \{ & \forall \alpha. f \langle \alpha \rightarrow int \rangle \\
          & \forall \alpha. f \langle nat \rightarrow \alpha \rangle \}
\end{align*}

@@

Therefore, the previous problem would never generate the clause \(f \langle nat \rightarrow int \rangle\).
[having trouble finishing the example, we would need for the nat int clause to allow finding a contradiction
without allowing the other two to generate this one]

@@

@@

\ForAll {clause \(C \in P\)}
   \ForAll {\(subst \in S(C)\)}
      \State \(C' = apply(subst, C)\)

      \If {\(C'\) is ground}

         \State \(P = P \cup \{C'\}\)
      \EndIf

   \EndFor
\EndFor

@@

@@

Comments on results:
So far from the testing i've done:
\begin{itemize}
   \item we gain very few problems compared to portfolio mode, this is very bad
   \item these can easily be lost with small bound adjustments, this isn't great but not catastriphic
   \item almost all newly gained problems are solved with the 0.0 e call, this one is catastrophic
\end{itemize}

@@

>>>>

TODO:
   \begin{itemize}
      \item \st{contact Sascha Boehme about authorship}

      \item it seems that dissentangling the implemntation and low-level description is quite difficult and may not be appropriate, it might be best to instead fuse low-level description and implementation and then split the section into substitution generation and clause generation

      \item are there any implementation of this (or comparable) algorithm in that have been evaluated?
      \item read mangling paper in detail
      \item finish implementation (mostly there)
      \item test and evaluate implementation (well on the way)
      \item (can wait) merge implementation (review code with active contributor(s) beforehand)
      \item find pretty inline setup for pseudo code
      \item (in order) write high/low level algo
      \item write implementation and evaluation
      \item write background
      \item write related work
      \item write conclusion
      \item write intro
      \item reread, spellcheck ...
    \end{itemize}

@@

\subsection{Soundness}
[this isn't an actual proof, more of a justification, mostly a joke in its present form]
Soundness of the algorithm follows from the fact that the resulting clauses are obtained by directly instantiating universally quantified type variable of the initial clauses. Then we rely upon the soundness of mangling [include reference].

\subsection{Multiple type arguments}
Comment that a function symbol with multiple type arguments will often see them instantiated one at a time per iteration. Conclude that it may be beneficial to remove clauses with function symbols with too many uninstantiated type arguments. Nuance that if the more there are such symbols, the more likely we are to generate substitutions that instantiate several type arguments at a time. Note that in practice setting a type variable limit changes nothing.
[more tests needed for showing that
[need to add a couple of tests to show that]
[maybe this section doesn't belong here, probably better in the low level section]

@@

This algorithm is incomplete, this is because not all possible instantiations of the type variables are generated.
Consider the following problem [note that we need a consistent notation for type arguments (notably to distinguish them from normal arguments) also this might not be an issue but they are terms not literals ]: 
\begin{align*}
   P = \{ & \forall \alpha. f \langle \alpha \rightarrow int \rangle \\
          & \forall \alpha. f \langle nat \rightarrow \alpha \rangle \\
          & \forall \alpha \beta. \lnot f \langle \alpha \rightarrow \beta \rangle \}
\end{align*}

@@

Therefore, the previous problem would never instantiate a clause, despite the problem becoming trvially contradictory if the the last clause and any of the first two are instantiated.

@@

\EndFunction
\end{algorithmic}
\end{algorithm}
In the zipperposition implemntation this iteration procedure is limited by several bounds. The number of new type arguments that can be generated for each function symbol is limited by user-defined bounds. These bounds can be defined proportionally. Additionally, the number of new type arguments can be limited at the clause level. In both cases, the bounds for the new ground type arguments and non-ground type arguments can be set independently.

A consequences of these bounds is that in order to avoid unecessary computations, the program will iteratively apply substitutions until either the ground or non-ground bound is reached. Then substitutions can be discarded depending on whether they instantiate or not a given set of type arguments

Additionally, there are bounds for the number of substitutions that are generated.

[So far, after more or less blind experimentation, I've settled on severely limiting new non-ground type arguments because they used to balloon out of control, whilst the growth in ground type arguments is more manageable]
[It's also worth suggesting the possibility of a more sophisticated implementation that could allow for dynamic bounds]

\begin{algorithm}
\begin{algorithmic}[1]
\Function{Instantiate clause}{Clause: C, Substitutions S}

   \If{\(C\) is ground}

      \State \Return \(\{C\}\)

      \State let \(v\) be a type variable in \(C\)
      \State let new\_clauses be an empty set of clauses

      \ForAll{subst \(\in S(C)\)}

         \If{subst instantiates \(v\)}

            \State \(C' = apply(subst, C)\)

            \State let v\_clauses = Instantiate clause(C', S)

         \EndIf

         \State \( new\_clauses = new\_clauses \cup v\_clauses\)

      \State \Return new\_clauses
   \EndIf

\EndFunction
\end{algorithmic}
\end{algorithm}
This function differs from the implementation in a significant way, indeed, through experimentation it was found that the generated clauses were far more likely to be solved by E if the substitutions that are combined to instantiate the clause were derived from the same iteration. In practice, we therefore store the iteration at which each substition is generated and at the instantiation phase only apply substitutions from the same iteration.

[need to make a handful of tests to check that]

What the previous function does is essentially going through the tree of possible substitutions that instantiate the clause and collecting all possible instantiations. This is extremely explosive, and a hard cap on the number of clauses we want to generate is almost indispensable for any implementation.

@@

\ForAll {clause \(C \in P\)}
   \State let new\_clauses = Instantiate clause(C, S)
   \State \(P = P \cup new\_clauses\)
\EndFor

@@

Provide complete list of problems explain which ones we excluded and why: for some it was lack of axiom files, a handful of others were simply too large

@@

Comments on results:
So far from the testing i've done:
\begin{itemize}
   \item we gain very few problems compared to portfolio mode, this is very bad especially since we only look at polymorphic problems, where this algorithm should shine at least in proportional terms
   \item these can easily be lost with small bound adjustments, this isn't great but not catastriphic
   \item almost all newly gained problems are solved with the 0.0 e call, this one is catastrophic
\end{itemize}

@@

% summary 
In conclusion, this paper has given a theoretical and practical description of an iterative monomorphisation algorithm. The practical description gives some insights on how it may be efficiently implemented. Additionally, its implementation in zipperposition demonstrates its practical feasibility.

However, the evaluation of this implementation against the TPTP benchmarks show dissapointing results in light of the efficiency of similar ideas in other frameworks. Indeed, zipperposition can only solve a handful of additional problems thanks to monomorphisation. Despite these issues, a refinment of the existing implementation paired with a better choice of bounds and options may allow for an improvement in results [decent enough to warrant having spent so much time on it]

% Future avenues
